---
title: Exploring sample overlap simulations
---


```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(simulateGP)
	library(TwoSampleMR)
	library(dplyr)
	library(ggplot2)
	library(parallel)
	library(knitr)
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
```

## Outline of simulation

1. Create x and y in a population
2. Find instruments for x in discovery and test for significance
3. Retest in replication
4. Extract those effects from outcome GWAS, which could be independent, discovery or replication sample



### Find GWAS parameters which will likely give high winner's curse


```{r}
simulate_gwasx <- function(nid, nsnp, bgx, bux)
{
	g <- make_geno(nid, nsnp, 0.5)
	u <- rnorm(nid)
	geneff <- rep(bgx, nsnp)
	x <- make_phen(c(geneff, bux), cbind(g, u))
	gwas(x, g)
}


sapply(1:100, function(x)
{
	simulate_gwasx(3000, 20, 0.08, 0.4) %>%
	{sum(.$pval < 5e-8)}
}) %>% table
```


### Check that UMVCUE works at reducing winner's curse

```{r}

umvcue <- function(d_beta, r_beta, d_se, r_se, pthresh)
{
	t <- qnorm(1 - pthresh/2)
	MLE <- (r_se^2 * d_beta + d_se^2 * r_beta) / (r_se^2 + d_se^2)
	BB <- (sqrt(r_se^2 + d_se^2) / d_se^2) * (MLE - t * d_se)
	b <- MLE - (r_se^2 / sqrt(r_se^2 + d_se^2) * dnorm(BB) / pnorm(BB))
	return(b)
}


simulate_umvcue <- function(nid, nsnp, bgx, bux)
{
	geneff <- rep(bgx, nsnp)

	g <- make_geno(nid/2, nsnp, 0.5)
	u <- rnorm(nid/2)
	x <- make_phen(c(geneff, bux), cbind(g, u))
	d <- gwas(x, g)
	g <- make_geno(nid/2, nsnp, 0.5)
	u <- rnorm(nid/2)
	x <- make_phen(c(geneff, bux), cbind(g, u))
	r <- gwas(x, g)
	index <- d$pval < 5e-8
	dat <- tibble(
		truth = geneff,
		disc = d$bhat,
		rep = r$bhat,
		disc_sig = d$pval < 5e-8,
		what="all"
	)
	dat$umvcue <- NA
	if(any(index))
	{
		dat$umvcue[index] <- umvcue(d$bhat[index], r$bhat[index], d$se[index], r$se[index], 5e-8)
	}
	return(dat)
}


out <- lapply(1:100, function(x)
{
	simulate_umvcue(3000, 20, 0.13, 0.4) %>%
	mutate(sim=x)
}) %>% bind_rows()

out2 <- group_by(out, sim) %>%
	summarise(
		nsig=sum(disc_sig), 
		truth=mean(truth),
		disc = mean(disc),
		rep = mean(rep),
		umvcue = mean(umvcue, na.rm=T)
	)

out3 <- filter(out, disc_sig) %>%
	group_by(sim) %>%
	summarise(
		nsig=sum(disc_sig), 
		truth=mean(truth),
		disc = mean(disc),
		rep = mean(rep),
		umvcue = mean(umvcue, na.rm=T)
	)

colMeans(out2)
colMeans(out3)

```

it does reduce winner's curse, gives simular result to replication


### Run MR simulations for different sampling strategies

```{r}
gwas_bias <- function(gwas)
{
	# sum((gwas$bhat - gwas$b)^2)
	mean(abs(gwas$bhat - gwas$b))
}


gwas_sim <- function(nid = 9000, nsnp = 20, bgx = 0.08, bxy = 0.2, buy = 0.4, bux = 0.4, out_index = 1:3000)
{
	# Simulate population
	g <- make_geno(nid, nsnp, 0.5)
	u <- rnorm(nid)
	geneff <- rep(bgx, nsnp)
	x <- make_phen(c(geneff, bux), cbind(g, u))
	y <- make_phen(c(buy, bxy), cbind(u, x))

	# Get observational estimate
	obs_beta <- summary(lm(y ~ x))$coefficients[2,1]

	# GWAS in samples
	disc_index <- 1:3000
	rep_index <- 3001:6000
	disc_gwas <- gwas(x[disc_index], g[disc_index,])
	rep_gwas <- gwas(x[rep_index], g[rep_index,])
	out_gwas <- gwas(y[out_index], g[out_index,])

	disc_gwas$b <- geneff
	rep_gwas$b <- geneff
	out_gwas$b <- geneff * bxy

	bias <- expand.grid(sample=c("disc", "rep", "out"), what=c("all", "sig"), bias=NA)
	bias$sse[1] <- gwas_bias(disc_gwas)
	bias$sse[2] <- gwas_bias(rep_gwas)
	bias$sse[3] <- gwas_bias(out_gwas)

	# what is significant
	index <- disc_gwas$pval < 5e-8
	bias$sse[4] <- gwas_bias(disc_gwas[index,])
	bias$sse[5] <- gwas_bias(rep_gwas[index,])
	bias$sse[6] <- gwas_bias(out_gwas[index,])
	bias$nsig <- sum(index)

	# Perform MR
	d <- list()
	d$disc_all <- simulateGP::make_dat(disc_gwas, out_gwas)
	d$rep_all <- simulateGP::make_dat(rep_gwas, out_gwas)
	d$disc_sig <- simulateGP::make_dat(disc_gwas[index,], out_gwas[index,])
	d$rep_sig <- simulateGP::make_dat(rep_gwas[index,], out_gwas[index,])
	d$umvcue_sig <- d$disc_sig
	d$umvcue_sig$beta.exposure <- umvcue(d$disc_sig$beta.exposure, d$rep_sig$beta.exposure, d$disc_sig$se.exposure, d$rep_sig$se.exposure, 5e-8)

	m <- lapply(names(d), function(x) {
		y <- d[[x]]
		if(nrow(y) > 0)
		{
			z <- mr(y, method_list=c("mr_ivw", "mr_wald_ratio")) %>% as_tibble()
			z$what <- x
			return(z)
		} else {
			return(NULL)
		}
	}) %>% bind_rows()

	return(list(bias=bias, mr=m))
}


res <- lapply(1:100, function(i) {
	message(i)
	bind_rows(
		gwas_sim(out_index = 1:3000)$mr %>% mutate(overlap="disc_overlap"),
		gwas_sim(out_index = 3001:6000)$mr %>% mutate(overlap="rep_overlap"),
		gwas_sim(out_index = 3001:6000)$mr %>% mutate(overlap="no_overlap"),
		gwas_sim(out_index = 1:6000)$mr %>% mutate(overlap="both_overlap")
	)
}) %>% bind_rows()

```

Create figure

```{r}
ress <- filter(res) %>%
	tidyr::separate(what, into=c("exposure_dataset", "instruments"), "_") %>%
	group_by(exposure_dataset, instruments, overlap) %>%
summarise(b=mean(b), n=n(), se=sd(b)/sqrt(n))


ggplot(ress, aes(x=overlap, y=b)) +
geom_bar(stat="identity", position="dodge", aes(fill=exposure_dataset)) +
facet_grid(. ~ instruments) +
scale_fill_brewer(type="qual")
```



To do:

1. Can we add monte carlo standard errors to the figure, i.e. does UMVCUE have smaller standard error than replication?
2. Better labelling



Conclusion - How to use this for MR?

- Winner’s curse is an issue
- Sample overlap is an issue
- Weak instrument bias is an issue

Recommendation - either do

- external discovery; ukbb replication; ukbb outcome
- ukbb discovery; ukbb replication; external outcome




To do:

1. We have used extreme simulation examples. What is a more realistic result? Can we do simulations (2) and (3) using realistic data
	- e.g. 8 million SNPs, 10000 causal variants, explaning 50% variance, 400k samples
	- either very slow simulations using current method or theoretical estimates not using individual level values using code that started here https://github.com/explodecomputer/simulateGP/blob/master/R/theoretical_gwas.R
2. Perform discovery and replication in UKBB for instruments
	- this is where the GWAS results are `/mnt/storage/private/mrcieu/research/scratch/IGD/data/public/UKB-b-*/clump.txt`. It includes the complete GWAS summary data on ~400k samples, plus the clumped top hits. The task is to
		1. Go back to the original 400k individual level data
		2. Split it into two - 200k and 200k (to do this you don't change the genetic data, you just set half the phenotype value to NA)
		3. Re-estimate the clumped effects in each of the datasets



